---
title: "Lecture 2 Code"
author: "Alex Hoagland"
date: "`r Sys.Date()`"
output: word_document
---

If you want to customize your .rmd file's look, you can add some options here. You can also load some packages here.

```{r setup, include=FALSE}
# Can add a header to your document if you want (this is a comment)

knitr::opts_chunk$set(echo = TRUE)

# Packages we'll need
# install.packages("chattr")
library(chattr)
library(tidyverse)
library(readxl)
# ?chattr

# Use your Chat-GPT Open AI API key here: https://platform.openai.com/api-keys
# Sys.setenv(OPENAI_API_KEY = "YOUR KEY HERE") 
# chattr_app()
```

## Variables

First, we can define some useful variables:

```{r variables}

# Basic variables
a <- 1
b <- 2

# Let's try out copilot! (This didn't help me)
# For each letter a, b, c, and d, assign a random number between 1 and 100

# Let's try out chattr! (This did help me)
# Load necessary libraries


# Create a data frame with a column 'letter' containing letters a to z
letters_df <- data.frame(letter = letters)

# Assign a random number between 1 and 1000 to each letter
set.seed(123) # for reproducibility
letters_df <- letters_df %>%
  mutate(random_number = sample(1:1000, n(), replace = TRUE))

letters_df
```

Notice that when you knit your file, this will "spit out" the table above. This is like circling your answer in math homework – you want to publish your work outside of the code so that we can see your final answer.

What about other variable types?

```{r variable types}
mychar <- "test"
mybool <- TRUE
# we'll cover factors more later. 

as.numeric(mybool)
as.numeric(mychar) # what does this mean? 
as.numeric(mychar == "test")

```

## Calculations in R 

Let's try some simple calculations – shout them out. Something to think about -- do we want to save them as objects or export them as output? 

```{r calculations}
a + b
```

## Now vectors

```{r vectors}
# Manual assignment
myvec <- c(1,3,45,6,0)
myvec2 <- 1:5
myvec + myvec2

# Evenly spaced vectors
myvec3 <- seq(from=10,to=1000,length=5)
myvec+myvec2+myvec3

# Random numbers
myvec4 <- runif(5, min=0, max=100)
myvec+myvec2+myvec3+myvec4
```
Do we like the way this output looks? What if you run the whole chunk at once? How can we do better?

Now let's summarize a vector here: 

```{r summary} 
outvec <- myvec + 3*myvec3

# What do each of these tell us? 
mean(outvec)
min(outvec)
median(outvec)
sd(outvec)
range(outvec)
length(outvec)
```

## What have we got so far? 

``` {r str}
str(a)
str(myvec3)
str(letters_df)
```

## Matrices
We can go from vectors to matrices easily. 

```{r matrices}
# Manual matrix
mymat <- matrix(c(1:15), nrow=5, ncol=3)
mymat 

# Combine myvec, myvec2, and myvec3 into a matrix (copilot)
mymat2 <- cbind(myvec, myvec2, myvec3)
mymat + mymat2

mymat * mymat2 # what is this doing? 
# mymat %*% mymat2 # Why doesn't this work? 
t(mymat) %*% mymat2 # What does this do? 

# Most of the time, we just want the data
mydata <- as.data.frame(mymat2)
# VieW(mydata)
```

## Now let's practice working with data

Let's load the dataset "Uninsured.xlsx" from our GitHub repo, which you will use for one of the upcoming assignments. 

```{r loaddata}
# Load the data
uninsured <- read_excel("C:\\Users\\alexh\\Dropbox\\Teaching\\HAD5772_Stats\\HAD5772_2025W\\HAD5772\\Datasets\\Uninsured.xlsx") 
# View(uninsured)

# Summmarize the data
names(uninsured)
mean(uninsured)
summary(uninsured)
uninsured %>% head()
str(uninsured)

# What is the average bankruptcy rate in areas with large populations? 
summary(uninsured$Population)
uninsured %>% filter(Population >= 80351) %>% summarize(avgbank = mean(frac_bankrupt))
uninsured %>% filter(Population < 80351) %>% summarize(avgbank = mean(frac_bankrupt)) # Do these seem different? We will learn how to test! 
```

## Visualizing data
Finally, let's visualize some data. 

```{r visualize}
# Let's make a histogram of the bankruptcy rate using ggplot (copilot)
ggplot(uninsured, aes(x=frac_bankrupt)) + geom_histogram(binwidth = 0.001, fill="blue", color="black") + labs(title="Bankruptcy Rate in US Counties", x="Fraction of Bankruptcies", y="Count")

# How can you customize this? 
```

Now this is where we will practice knitting our file! (The binscatter code segment below is for the second half of the lecture)

```{r binscatter}
# What does a raw scatterplot look like 
ggplot(uninsured, aes(x=Population, y=frac_bankrupt)) + geom_point()

# Create a binned scatterplot between bankruptcy rate and population using ggplot
uninsured %>%
  mutate(bin = ntile(Population, 5)) %>%
  group_by(bin) %>% summarize(xmean = mean(Population), ymean = mean(frac_bankrupt)) %>% 
  ggplot(aes(x = xmean, y = ymean)) +
  geom_point() +
  labs(title = "Bankruptcy Rate in US Counties by Population", x = "Population", y = "Fraction of Bankruptcies")
```

What stories do you see here? Are they robust (to what)?

