---
title: "Lecture 3 Code"
author: "Alex Hoagland"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Here is my header text, where I load my packages and look at documentation for them: 

```{r header}
# I like to include several additional notes in the header of my files here:
#
# Last modified: 1/20/2025
#
### PURPOSE:
  # Lecture 3 code and output file
#
### NOTES:
  # - uses the Tidyverse package and Dplyr
  # - uses the NHANES package to load data from the US National Health and Nutrition Examination Survey (NHANES, 1999-2004).

library(tidyverse) # load the installed package for each new session of R
library(NHANES) # where we'll get our data extract
?NHANES # documentation for a package or function looks like this! This one gives us a list of all variables we could want
```

## Loading Data

Let's load some data from the NHANES and save it as our own object.

```{r load-my-data}
mydata <- NHANES # gives us a random smaple of 10,000 observations
  # note from documentation: simple random sample of the American population
View(mydata) # Let's look at the structure
  # What does each row (observation) represent? What kinds of variables do we have? How would we describe this dataset in words? 
```

Now how would we describe this dataset in numbers? 

## Simple tables and figures

Remember that tables are generally easier to work with outside of knitting, but you might want some simple code to generate the numbers

```{r tables-figures}
# Let's summarize days of poor mental health by education 
levels(mydata$Education) # What does this tell us? 
tabledata <- mydata %>% group_by(Education) %>% 
  summarize(mean_days_mental_health = mean(DaysMentHlthBad, na.rm=T),
            median_days_mental_health = median(DaysMentHlthBad, na.rm=T),
            n = n()) # Note that this came from copilot! 
kableExtra::kable(tabledata) # Now you can format this in Word! This is what I recommend for the next assignment. 

# simple histogram
barplot(prop.table(table(mydata$Education))) # What does this tell us? 
barplot(prop.table(table(mydata$Education,mydata$DaysMentHlthBad))) # What does this tell us? 
```

## Histograms
Now let's start visualizing the data.

```{r visualize-continuous-vars}
# first, a histogram of our two variables
ggplot(data=mydata, aes(x=Age)) + geom_histogram() # the simplest version. How can we clean it up? Let's do this live.

# A final version
# includes: bin width, color, labels, theme
ggplot(data=mydata, aes(x=Age)) + geom_histogram(binwidth = 5, color="gray", fill="blue", alpha=.6) + labs(y = "") + theme_minimal()

# A note on bin widths: what do you notice about these? What is similar/different?
ggplot(data=mydata, aes(x=Age)) + geom_histogram(binwidth = 2, color="gray", fill="blue", alpha=.6) + labs(y = "") + theme_minimal()
ggplot(data=mydata, aes(x=Age)) + geom_histogram(binwidth = 5, color="gray", fill="blue", alpha=.6) + labs(y = "") + theme_minimal()

# with enough data, can we ignore bin widths altogether?
ggplot(data=mydata, aes(x=Age)) + geom_density(color="black", fill="blue", alpha=.6) + labs(y = "") + theme_minimal()

# Now let's look at Days of Poor Mental Health
ggplot(data=mydata, aes(x=DaysMentHlthBad)) + geom_histogram(binwidth = 3, color="gray", fill="red", alpha=.6) + labs(y = "") + theme_minimal()
```

## Measures of central tendency
The histogram provides a nice overview, but we need some numbers to summarize lots of data in a simple way. 

What if we try to summarize this Days of Poor Mental Health variable? 

```{r mct}
mean(mydata$DaysMentHlthBad) # uh oh! 
mean(mydata$DaysMentHlthBad, na.rm=T) # what did we do here? 

median(mydata$DaysMentHlthBad, na.rm=T) # what do we conclude? 
quantile(mydata$DaysMentHlthBad, 
         probs = c(0.1, 0.25, 0.5, 0.75, 0.9), 
         na.rm=T) # what does this tell us? 

# What if we want to zero in on a particular group? 
subset <- mydata %>% filter(DaysMentHlthBad >= 1) # What did we do? 
quantile(subset$DaysMentHlthBad, 
         probs = c(0.1, 0.25, 0.5, 0.75, 0.9), 
         na.rm=T) # what does this tell us? 

# A (bit of a) shortcut
summary(mydata$Age) # What do we see?
summary(mydata$DaysMentHlthBad) # Note the NAs, what does this mean?
```

Now what about variation? 

```{r variation}
sd(mydata$DaysMentHlthBad, na.rm=T) # How do we measure this? 
  # This tells us something about where the "useful" bit of data comes in
var(mydata$DaysMentHlthBad, na.rm=T) # what is this in relation to SD?
sd(subset$DaysMentHlthBad, na.rm=T) # Why is this larger than before? 

# This adds SDs to our histogram to show that we capture most of our data within a standard deviation
mymean <- mean(mydata$DaysMentHlthBad, na.rm = T)
mysd <- sd(mydata$DaysMentHlthBad, na.rm = T) # What does this mean?

mydata <- mydata %>% drop_na(DaysMentHlthBad) %>%
  mutate(myfill = ifelse(abs(DaysMentHlthBad-mymean)<=mysd*2, 1, 0))
mydata$myfill <- factor(mydata$myfill)
ggplot(mydata, aes(x=DaysMentHlthBad, fill = myfill)) +
  geom_histogram(width = 1, color='gray') +
  labs(y = "", fill = "Within 1 SD") + theme_minimal()
```
```

## Summarizing and Visualizing Data

For most of our data visualization needs, we will use ggplot. Note that data visualization is extremely important! Captivating visualizations make a great paper stand out, and bad ones can sink even the best paper.GGplot is extremely versatile! I recommend this book if you are looking to master ggplot: <https://tinyurl.com/4k4wj8px>

Let's start with two continuous variables: age and days of poor mental health. We'll do a simple scatterplot and then a binned scatterplot: 

```{r scatterplot}
# Simple scatterplot between age and DaysMentHlthBad
ggplot(data=mydata, aes(x=Age, y=DaysMentHlthBad)) + geom_point() # What does this tell us?

# Now let's create a binned scatterplot 
mydata %>%
  mutate(bin = ntile(Age, 100)) %>% 
    # What should we play around with to make this sensible? 
  group_by(bin) %>% 
  summarize(xmean = mean(Age, na.rm=T), 
            ymean = mean(DaysMentHlthBad, na.rm=T)) %>% 
  ggplot(aes(x = xmean, y = ymean)) +
  geom_point() + theme_minimal() + 
  geom_vline(xintercept = 18) + geom_vline(xintercept = 55) + 
  labs(x = "Respondent Age", y = "Average Days with Poor Mental Health")
```

### Covariance and Correlation

```{r covariance-correlation}
cov(mydata$DaysMentHlthBad, mydata$Age,use="pairwise.complete.obs") # Base covariance -- but in units of what?
cov(mydata$DaysMentHlthBad/7, mydata$Age,use="pairwise.complete.obs")

cor(mydata$DaysMentHlthBad, mydata$Age,use="pairwise.complete.obs") # This makes more sense!
cor(mydata$DaysMentHlthBad/7, mydata$Age,use="pairwise.complete.obs") # This makes more sense!

# What makes this number useful for you? What would you think about changing to make this number more useful, given our results above? 
```
